Data Descriptor for Nature Scientific Data

### 2024-10-29 manuscript/manuscript-nature/manuscript-nature.qmd
headlines-and-writeup-results-part1

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.

![Classification of spatial units into categories by population density and by relative deprivation index. Spatial distribution of categories and population share in each category, by country.](figures/Fig1.png){#Fig1}

## COVID-19’s impact on human mobility displays consistently uneven patterns across population groups in Latin American countries

The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.
Figure @Fig2 shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
These changes show remarkably consistent patterns across countries for both types of classification.

Focusing first on the early pandemic days, we observe a decrease in the number of inflows for all countries and most categories of analysis.
This decrease displays a systematic gradient across categories of analysis whereby the more densely populated or the more affluent an area is, the higher the losses in the number of inflows.
Differences in the levels of mobility across population density or RDI categories tend to diminish with time, with a less pronounced gradient across categories of analysis approximately two years later.

These patterns can be attributed to...
remote work, movement to rural areas, affluent people being more able to abide to lockdowns, etc...
and the patterns are consistent with what has been found in ...

![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig2.png)

## Recovery trajectories towards baseline mobility levels are heterogeneous across population categories in Latin America

In this section we explore trends in the intensity of movements over time across population groups.
Generally, all categories of analysis experienced an initial drop in the number of movements followed by an upwards general trend of recovery towards baseline levels of mobility.
The extent of the initial drop and the speed of recovery varied by country and by analysis category.
These general patterns can be visually explored in Figure @Fig3 and @Fig4 for a classification by population density category and by RDI respectively.
Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.

![](figures/Fig3.png)

![](figures/Fig4.png)

@@

# Data and methods {.unnumbered}

@@

@@

>>>>

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America. We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.

The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.

## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels

We first focus the analysis on the variability in COVID-19's impact on daily mobility levels across different population groups. Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022. These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset. May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures. March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis. Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index. The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$. Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction. These changes show remarkably consistent patterns across countries for both types of classification.

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days. The extent of this decrease tends to be larger for high-density areas. Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density. In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%. However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes. Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions. In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories. Notably, mobility levels in areas characterised by higher population density still remain below the baseline. Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline. These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May). Overall, the observed patterns by population density category are consistent with those reported in a previous work CITE, where the same data was analysed without applying the data processing methodology proposed in this study.

Moving to the patterns by relative deprivation index category, we observe tendencies that mirror the patterns by population density category. Firstly, in March 2020, there is an overall decrease in the number of inflows for all countries and most categories of analysis. Exceptions to this general tendency notably occur in the most deprived (less affluent) areas, especially in Colombia, where mobility levels in some regions are actually above the baseline. This decrease displays a gradient across categories of analysis whereby the more affluent an area is, the higher the losses in the number of inflows. After almost two years, we find evidence for a general recovery of mobility closer to baseline levels across groups regardless of how deprived they are. However, disparities between groups are still visible, indicating that the pandemic has exacerbated behavioral differences among groups with varying socioeconomic statuses.

These patterns can be attributed to...
THIS GOES TO THE DISCUSSION

![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig-impact.png)

## Initial drop in mobility drives current levels, as recovery pace remains consistent across population groups

Next we focus on

In this section we explore trends in the intensity of movements over time across population groups.
Generally, all categories of analysis experienced an initial drop in the number of movements followed by an upwards general trend of recovery towards baseline levels of mobility.
The extent of the initial drop and the speed of recovery varied by country and by analysis category.
These general patterns can be visually explored in Figure @Fig-recovery-density and @Fig-recovery-rdi for a classification by population density category and by RDI respectively.
Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.

![](figures/Fig-recovery-density.png)

![](figures/Fig-recovery-rdi.png)
![](figures/Fig-re-density.pdf)

![](figures/Fig-re-rdi.pdf)

@@

# Data {.unnumbered}

@@

# Methods

@@

The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.

![Classification of spatial units into categories by population density and by relative deprivation index. Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#Fig1}
